---
title: "Getting Started with nycOpenData"
output: rmarkdown::html_vignette
author: "Christian Martinez"
vignette: >
  %\VignetteIndexEntry{Getting Started with nycOpenData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(nycOpenData)
library(ggplot2)
library(dplyr)
```

## Introduction

Welcome to the `nycOpenData` package, a R package dedicated to helping R users connect to the [NYC Open Data Portal](https://opendata.cityofnewyork.us/)!

The `nycOpenData` package provides a streamlined interface for accessing New York City's vast open data resources. It connects directly to the NYC Open Data Portal, helping users bridge the gap between raw city APIs and tidy data analysis. It does this in two ways:

### The `nyc_pull_dataset()` function

The primary way to pull data in this package is the `nyc_pull_dataset()` function, which works in tandem with `nyc_list_datasets()`. You do not need to know anything about API keys or authentication.

The first step would be to call the `nyc_list_datasets()` to see what datasets are in the list and available to use in the `nyc_pull_dataset()` function.

```{r nyc-list-datasets}
nyc_list_datasets()
```

The output includes columns such as the dataset title, description, and link to the source. The most important piece is the key, because the key is what's needed for the `nyc_pull_dataset()` function.

```{r nyc-311-pull}
nyc_311_data <- nyc_pull_dataset(key = "nyc_311", limit = 3)
```

Now, we have the 3 most recent 311 requests!

### The `nyc_any_dataset()` function

The easiest workflow is to use `nyc_list_datasets()` together with `nyc_pull_dataset()`. However, there are ample datasets on the portal, with new ones being added all the time, and so the list does not have all of the datasets.

In the event that you have a particular dataset you want to use in R that is not in the list, you can use the `nyc_any_dataset()`. The only requirement is the dataset’s API endpoint (a URL provided by the NYC Open Data portal). Here are the steps to get it:

1. On the NYC Open Data Portal, go to the dataset you want to work with.
2. Click on "Export" (next to the actions button on the right hand side).
3. Click on "API Endpoint".
4. Click on "SODA2" for "Version".
4. Copy the API Endpoint.

Below is an example of how to use the `nyc_any_dataset()` once the API endpoint has been discovered, that will pull the same data as the `nyc_pull_dataset()` example:

```{text}
nyc_311_data <- nyc_any_dataset(json_link = "https://data.cityofnewyork.us/resource/erm2-nwe9.json", limit = 3)
```

### Rule of Thumb

While both functions provide access to NYC Open Data, they serve slightly different purposes.

In general:
- Use `nyc_pull_dataset()` when the dataset is available in `nyc_list_datasets()`
- Use `nyc_any_dataset()` when working with datasets outside the built-in catalog

Together, these functions allow users to either quickly access curated datasets or flexibly query any dataset available on the NYC Open Data portal.

## Real World Example

NYC has a population of almost 8.5 million people. By calling 311, residents are able to make comments, inquiries, complaints, and requests to the city agencies. All 311 service requests are contained in the dataset, [found here](https://data.cityofnewyork.us/Social-Services/311-Service-Requests-from-2020-to-Present/erm2-nwe9/about_data). In R, the `nycOpenData` package can be used to pull this data directly.

By using the `nyc_pull_dataset()` function, we can gather the most recent 311 calls in New York City, and filter based upon any of the columns inside the dataset.

### Pulling a Small Sample

To start, let's pull a small sample to see what the data looks like. By default, the function pulls in the *10,000 most recent* requests, however, let's change that to only see the latest 3 requests. To do this, we can set `limit = 3`.

```{r small-sample}
small_sample <- nyc_pull_dataset("nyc_311", limit = 3)
small_sample

# Seeing what columns are in the dataset
names(small_sample)
```

Fantastic! We successfully pulled 311 data from the NYC Open Data Portal.

Let's now take an example of the last 3 requests from the borough Brooklyn. The `nyc_311()` function can filter based off any of the columns in the dataset. To filter, we add `filters = list()` and put whatever filters we would like inside. From our `colnames` call before, we know that there is a column called "borough" which we can use to accomplish this.

```{r filter-brooklyn}

brooklyn_311 <- nyc_pull_dataset(key = "nyc_311",limit = 3, filters = list(borough = "BROOKLYN"))
brooklyn_311

# Checking to see the filtering worked
brooklyn_311 |>
  distinct(borough)
```

Success! From calling the `brooklyn_311` dataset we see there are only 3 rows of data, and from the `distinct()` call we see the only borough featured in our dataset is BROOKLYN.

One of the strongest qualities this function has is its ability to filter based off of multiple columns. Let's put everything together and get a dataset of the last *50* 311 requests from the New York Police Department in Brooklyn.

```{r filter-brooklyn-nypd}
# Creating the dataset
brooklyn_nypd <- nyc_pull_dataset("nyc_311", limit = 50, filters = list(agency = "NYPD", borough = "BROOKLYN"))

# Calling head of our new dataset
brooklyn_nypd |>
  slice_head(n = 6)

# Quick check to make sure our filtering worked
brooklyn_nypd |>
  summarize(rows = n())

brooklyn_nypd |>
  distinct(agency)

brooklyn_nypd |>
  distinct(borough)
```

We successfully created our dataset that contains the 50 most recent requests regarding the NYPD in the borough Brooklyn.

### Mini analysis

Now that we have successfully pulled the data and have it in R, let's do a mini analysis on using the `complaint_type` column, to figure out what NYC residents in Brooklyn are complaining about to the NYPD.

To do this, we will create a bar graph of the complaint types.

```{r compaint-type-graph, fig.alt="Bar chart showing the frequency of NYPD-related 311 complaint types in Brooklyn from the 50 most recent service requests.", fig.cap="Bar chart showing the frequency of NYPD-related 311 complaint types in Brooklyn from the 50 most recent service requests.", fig.height=5, fig.width=7}
# Visualizing the distribution, ordered by frequency
brooklyn_nypd |>
  count(complaint_type) |>
  ggplot(aes(
    x = n,
    y = reorder(complaint_type, n)
  )) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Most Recent NYPD 311 Complaints (Brooklyn)",
    subtitle = "Top 50 service requests",
    x = "Number of Complaints",
    y = "Type of Complaint"
  )
```

This graph shows us not only *which* complaints were made, but *how many* of each complaint were made.

## Summary

The `nycOpenData` package serves as a robust interface for the NYC Open Data portal, streamlining the path from raw city APIs to actionable insights. By abstracting the complexities of data acquisition—such as pagination, type-casting, and complex filtering—it allows users to focus on analysis rather than data engineering.

As demonstrated in this vignette, the package provides a seamless workflow for targeted data retrieval, automated filtering, and rapid visualization.

## How to Cite

If you use this package for research or educational purposes, please cite it as follows:

Martinez C (2026). nycOpenData: Convenient Access to NYC Open Data API Endpoints. R package version 0.2.0, <https://martinezc1.github.io/nycOpenData/>.
